WSClean: a new implementation of a fast, generic wide-field imager

### 2014-01-12 articles/2014/Article-WSClean/wsclean.tex

\documentclass[a4paper,10pt]{article} 
%\usepackage[style=nature,backend=biber]{biblatex}
\usepackage{graphicx}
\usepackage[font={it},labelfont={bf}]{caption}
\usepackage{times}
\usepackage{color}
%\usepackage{fixltx2e} % fixes placing of image positions
\usepackage{authblk}
\usepackage{amsmath}
\usepackage{courier}
\usepackage[round]{natbib}
\usepackage{hhline}

\definecolor{customhdrcolor}{rgb}{0.0,0.0,0.0}
%\definecolor{customcitecolor}{rgb}{0.0,0.25,0.25}
\definecolor{customcitecolor}{rgb}{0.0,0.5,0.75}
%\definecolor{customlinkcolor}{rgb}{0.0,0.0,1.0}
\definecolor{customlinkcolor}{rgb}{0.0,0.5,0.75}

@@

\ifpdf\pdfinfo{/Title      (MWA paper proposal Offringa)
               /Author     (A. R. Offringa)
               /Keywords   (MWA, imaging)
        }
\else\usepackage{graphics}\fi

@@

%\title{MWA Paper proposal:\\````WSClean: a fast implementation of a generic wide-field imager''}

%\author{A.~R.~Offringa}
%\author[1,2]{A.~R.~Offringa}
%\affil[1]{RSAA, Australian National University, Mt Stromlo Observatory, via Cotter Road, Weston, ACT 2611, Australia}
%\affil[2]{ARC Centre of Excellence for All-Sky Astrophysics (CAASTRO)}
\begin{document}

\label{firstpage}
\section*{MWA paper proposal: ``WSClean: a new implementation of a fast, generic wide-field imager''}
\begin{tabular}{ll}
Publication title: & WSClean: a new implementation of a fast, generic \\
                   & wide-field imager\\
                   & \\
Principle contact: & Dr. Andr\'e R. Offringa \\
                   & \\
Draft publication        & A.~R.~Offringa, B.~McKinley, N.~Hurley-Walker, \\
\hspace{5mm}author list: & F.~H.~Briggs, J.~Rhee, L.~Feng, D.~Kaplan, \\
                         & A.~R.~Neben, J.~D.~Hughes, MWA commissioning \\
                         & team and GLEAM members where appropriate, \\
                         & MWA builders list \\
                   & \\
Journal:           & Probably MNRAS\\
                   & \\
Intended date of submission\\
\hspace{5mm}to collaboration: & February 2014\\
\end{tabular}

\section*{Paper summary}
We present our new imager implementation, ``WSClean'', which we find to be an order of magnitude faster than existing implementations, as well as being capable of MWA full-sky imaging at full resolution.

Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with simple imagers that are based on a two-dimensional Fast Fourier Transformation (FFT). Instead, the imaging algorithm needs to account for the ``w-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane. This deviation is amplified for telescopes with wide field of view, making this problem especially hard for low-frequency wide-field telescopes.

@@

After describing the new imaging implementation, we will test its performance using various MWA observations, including observations with varying elevation and fields (diffuse emission, point sources). GMRT data and -- if available -- LOFAR data will also be used. We will compare these results with the results from the w-projection implementation in CASA and to other non-w-projection algorithms, thereby focusing on accuracy and performance. The data sets that we currently envision to use are summarized in Table~\ref{tbl:observations}.

Once the paper is accepted for publication in a scientific journal, the WSClean source code will be publicly released.

\begin{figure*}[bh]
\begin{center}
\includegraphics[width=12cm]{EoR0-apparent.png}
\caption{A deep MWA image of EoR0 created from one hour of observing, imaged and deconvolved with WSClean in about an hour using 10 nodes on Fornax. Image has not been beam corrected. Noise level after correcting is 7 mJy. 200 sources were subtracted from the centre before imaging.}
\label{fig:stddev-spectrum}
\end{center}
\end{figure*}

\begin{table}
\caption{Data sets envisioned to be used in publication} \label{tbl:observations}
 \begin{tabular}{|c|l|p{6cm}|}
 \hhline{|=|=|=|}
 \textbf{Observatory} & \textbf{Description} & \textbf{Reasons} \\
 \textbf{+ date} & & \\
 \hhline{|=|=|=|}
  2013-05-02 & Decl -55          & $\bullet$ Test imaging at different frequencies\\
  MWA 128T   & GLEAM test data   & $\bullet$ Show computational performance \\
             &                   & \hspace{5mm} at elevation $\sim 60$\\
             &                   & $\bullet$ Full-sky imaging \\
 \hline
  2013-05-16 & Vela and Puppis A & $\bullet$ Test cleaning of diffuse structure \\
  MWA 128T   &                   & $\bullet$ Show computational performance \\
             &                   & \hspace{5mm} of cleaning\\
             &                   & $\bullet$ Pointed observations \\
 \hline
  2013-07-31 & EoR0              & $\bullet$ Test cleaning of point sources \\
  MWA 128T   &                   & $\bullet$ Test deep imaging \\
 \hline
  2008-2009  & ZCosmos           & $\bullet$ Test relevance to GMRT data \\
  GMRT       & (stacking)        & $\bullet$ Verify making image cubes \\
 \hline
  LOFAR      & To be determined  & $\bullet$ Test relevance to LOFAR data \\
             &                   & (only if data easily obtainable) \\
 \hline
             & Simulations       & $\bullet$ Validate photometry / astrometry \\
   ---       &                   & $\bullet$ Test computational performance \\
             &                   & \hspace{5mm} at very low elevation\\
 \hhline{|=|=|=|}
\end{tabular}
\end{table}

@@

\bibliographystyle{plainnat}
\bibliography{references}

>>>>

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage[font={it},labelfont={bf}]{caption}
%\usepackage{units}
\usepackage{times}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

\usepackage{fixltx2e} % fixes placing of image positions

\definecolor{customhdrcolor}{rgb}{0.0,0.0,0.0}
\definecolor{customcitecolor}{rgb}{0.0,0.5,0.75}
\definecolor{customlinkcolor}{rgb}{0.0,0.5,0.75}

@@

\ifpdf\pdfinfo{/Title      (WSClean)
               /Author     (A. R. Offringa et al.)
               /Keywords   (instrumentation: interferometers;methods: observational;techniques: interferometric;radio continuum: general)
        }
\else\usepackage{graphics}\fi

@@

%\newcommand{\editmark}[1]{#1}
%\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}

% To make Dutch ``tussenvoegsels'' work correctly in Latex such as ``de Bruyn'', we use this command
% It fixes ordering and uppercases.
% In text, it should be written with uppercase, as in ``De Bruyn''
\DeclareRobustCommand{\TUSSEN}[3]{#2}

\title[WSClean: a new imager]{WSClean: a new implementation of a fast, generic wide-field imager}

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{E-mail:
\url{andre.offringa@anu.edu.au}}, B.~McKinley$^{1,2}$, N.~Hurley-Walker, 
F.~H.~Briggs, \newauthor
J.~Rhee, L.~Feng, D.~Kaplan, A.~R.~Neben, J.~D.~Hughes, MWA commissioning team\newauthor
and GLEAM members where appropriate \& MWA builders list
\\
$^{1}$RSAA, Australian National University, Mt Stromlo Observatory, via Cotter Road, Weston, ACT 2611, Australia \\
$^{2}$ARC Centre of Excellence for All-sky Astrophysics (CAASTRO) \\
}

\begin{document}

\date{Accepted TODO. Received TODO; in original form TODO}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\pubyear{2014}

\label{firstpage}
\maketitle

\begin{abstract}
We present our new imager implementation, ``WSClean'', which we find to be an order of magnitude faster than existing implementations, as well as being capable of MWA full-sky imaging at full resolution.
TODO
\end{abstract}

\begin{keywords}
instrumentation: interferometers -- methods: observational -- techniques: interferometric -- radio continuum: general
\end{keywords}

\section{Introduction}

\section*{Acknowledgments}
...

% To make Dutch ``tussenvoegsels'' work correctly in Latex such as ``de Bruyn'', we use this command
% In bibliography, it should be written with lowercase, as in ``de Bruyn''
\DeclareRobustCommand{\TUSSEN}[3]{#3}

Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with simple imagers that are based on a two-dimensional Fast Fourier Transformation (FFT). Instead, the imaging algorithm needs to account for the ``w-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane. This deviation is amplified for telescopes with wide field of view, making this especially an issue for low-frequency telescopes, that by nature are wide-field instruments.

@@

After describing the new imaging implementation, we will test its performance using various MWA observations, including observations with varying elevation and fields (diffuse emission, point sources). GMRT data and -- if available -- LOFAR data will also be used. We will compare these results with the results from the w-projection implementation in CASA and to other non-w-projection algorithms, thereby focusing on accuracy and performance.

\section{The w-stacking technique}
In this section, we will describe the w-stacking algorithm from a mathematical point of view.

The sky equation for an interferometer is given by
\begin{eqnarray}\label{eq:visibility-function}
V(u,v,w) = \iint \frac{A(l,m) I(l,m)}{\sqrt{1-l^2-m^2}} e^{-2\pi i \left(ul + vm + w(\sqrt{1-l^2-m^2}-1)\right)} dl dm,
\end{eqnarray}
where function $V$ is formed by the calibrated visibilities, $u,v,w$ is a coordinate giving the observed mode in the coordinate system of the array, $A$ is the primary-beam function, $I$ is the sky function and $l,m$ are cosine sky coordinates. Imaging consists of inverting this equation to find $I$ from $V$.

For small field of views, the term $\sqrt{1-l^2-m^2}$ is approximately of unit size, transforming Eq.~\eqref{eq:visibility-function} into an ordinary 2-dimensional Fourier transform which can be inverted with an inverse Fast-Fourier Transform (FFT). A common rule is that this is valid when $\forall w,l,m: w\left(\sqrt{1-l^2-m^2}-1\right) \ll 1$.

TODO: Algorithm; plain w-stacking and further alterations such as w-snapshotting and full-sky imaging.
\section{The WSClean imager implementation}

\section{Analysis}
TODO: Show image quality; show time-efficiency comparisons.

\bibliographystyle{mn2e}
\bibliography{references}

@@
